---
title: "Mechanism-of-action Landscape"
author: 
  - name: Belinda B. Garana
package: MOAL
output:
  BiocStyle::html_document:
  vignette: >
    %\VignetteIndexEntry{Mechanism-of-action Landscape}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIEncoding{UTF-8}
      toc_float: true
      toc_depth: 2
---

# Introduction
Resistance is a major barrier to cancer remission. However, it is 
not practical to test all the millions of potential drug combinations 
due to cost limitations. To help prioritize drug pairings for further 
evaluation, we studied the similarity of drugs grouped by mechanisms-of-action 
(MOA) in published drug screens. With this process, termed the 
Mechanism-of-action Landscape (MOAL), 131 drug MOA pairings were prioritized
from more than 1.8 million potential drug pairings. Though this method does not 
evaluate synergy, it has been used to study more than a thousand drugs whereas 
other methods have only studied on the order of ten to one hundred drugs. In 
addition to recovering established drug combinations (e.g., MEK and RAF 
inhibitors in skin cancer), this method has also detected potential drug 
pairings which have not been well studied and demonstrated that results can 
vary depending on tissue type. Furthermore, we show that the MOAL method is 
beneficial over a simple UMAP visualization because it uses statistics to 
eliminate more than 96% of potential drug MOA pairings from consideration and 
identifies significant MOA pairings which may not be obvious from their 
distance on a UMAP visualization (e.g., BET and HDAC inhibitors). This method 
can also be used to evaluate the global performance of drug sensitivity 
prediction algorithms. For more information, please visit our website: 
https://belindabgarana.github.io/MOAL

# Installation
To install this package, start R and enter:
```{r eval=FALSE}
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("MOAL")
```

Alternatively, to install this package directly from GitHub, start R and enter:
```{r eval=FALSE}
if (!require("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

devtools::install_github("BelindaBGarana/MOAL")
```

If you are using Windows OS, you may need to change the code above to:
```{r eval=FALSE}
if (!require("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

devtools::install_github("BelindaBGarana/MOAL", build = FALSE)
```

# Input Option 1: Drug Screen Data Frame (<1k Drugs)
With the landscape function, you can rank drug moa's based on similarity to 
each other. For example, if your input data frame has sample names (e.g., 
CCLE IDs) in the first column and drug sensitivity scores (e.g., AUC) for 
various drugs in the rest of the columns, then moa pairs with NES_avg > 0 
may be toxic to the same samples and pairs with NES_avg < 0 may be toxic to 
opposite samples. In other words, if the moa pair "moa X & moa Y" passes 
quality control (qc) and NES_avg < 0, then drugs in moa X may be toxic to 
samples resistant to Y and vice-versa. The only requirements are that:
  * At least 2 drug sets are represented in your drug sensitivity data frame
  * At least 3 samples (e.g., cell lines) are in the drug sensitivity data frame

Required inputs:
  * drug.sensitivity: Data frame containing drug sensitivity scores (e.g., 
                      AUC) with drug names as column names, except for the 
                      first column which contain sample names (e.g., CCLE IDs)
  * drug.moa: Dataframe with moa annotations for each drug. 
              
Example:
1. Prepare drug.sensitivity data frame
```{r}
# create list of sample names
Sample_ID <- seq(from = 1, to = 21)

drug.sensitivity <- as.data.frame(Sample_ID)

# create list of drug names
Drug <- paste0("Drug_", seq(from = 1, to = 12))

# give each drug values representative of AUC sensitivity scores
for(i in seq_len(length(Drug))){
  drug.sensitivity[,c(Drug[i])] <- rnorm(length(Sample_ID),
                                        mean = 0.83, sd = 0.166)
}
```

2. Prepare drug.moa data frame with moa annotations for each drug
```{r}
# Drug is our default name for the column containing drug names
drug.moa <- as.data.frame(Drug)

# moa is our default name for the column containing drug moa set annotations
drug.moa$moa <- rep(paste("Set", LETTERS[seq(from = 1, to = 2)]), 6)
```

3. Perform moa landscape and store results
```{r}
landscape.test <- MOAL::landscape(drug.sensitivity, drug.moa)
```

# Input Option 2: Drug Screen Data Frame (>=1k Drugs)
If you are analyzing a large drug screen (>= 1k drugs), you may run out of
memory if you perform the full landscape analysis all at once. To prevent 
this issue, it is recommended to perform each of the 2 parts of the 
landscape analysis separately and clear your R environment after running 
the first part of the analysis (i.e., landscape_1d). Again, the only 
requirements are that:
  * At least 2 drug sets are represented in your drug sensitivity data frame
  * At least 3 samples (e.g., cell lines) are in the drug sensitivity data frame

Required inputs:
  * drug.sensitivity: Data frame containing drug sensitivity scores (e.g., 
                      AUC) with drug names as column names, except for the 
                      first column which contain sample names (e.g., CCLE IDs)
  * drug.moa: Dataframe with moa annotations for each drug. 
              
Example:
1. Prepare drug.sensitivity data frame
```{r}
# create list of sample names
Sample_ID <- seq(from = 1, to = 21)

drug.sensitivity <- as.data.frame(Sample_ID)

# create list of drug names
Drug <- paste0("Drug_", seq(from = 1, to = 12))

# give each drug values representative of AUC sensitivity scores
for(i in seq_len(length(Drug))){
  drug.sensitivity[,c(Drug[i])] <- rnorm(length(Sample_ID),
                                        mean = 0.83, sd = 0.166)
}
```

2. Prepare drug.moa data frame with moa annotations for each drug
```{r}
# Drug is our default name for the column containing drug names
drug.moa <- as.data.frame(Drug)

# moa is our default name for the column containing drug moa set annotations
drug.moa$moa <- rep(paste("Set", LETTERS[seq(from = 1, to = 2)]), 6)
```

3. Perform drug-level landscape, store results, and 
clear unnecessary data before continuing
```{r}
landscape.1d.test <- MOAL::landscape_1d(drug.sensitivity, drug.moa)

# after storing results, clear unnecessary data to free memory
landscape.1d.test <- landscape.1d.test$DMEA.results
```

4. Perform final step of MOA Landscape and store results
```{r}
landscape.2d.test <- MOAL::landscape_2d(landscape.1d.test)
```


# Input Option 3: Prediction Results
The MOA Landscape can also be used to evaluate the network-level performance
of drug sensitivity prediction algorithms. If you use one of these prediction 
algorithms to generate a predicted drug.sensitivity data frame, you can use 
that data frame as an input to the MOA Landscape (i.e., using one of the input 
options detailed above) and compare the landscape results to that of the 
original drug screen used to generate the prediction. Again, the only 
requirements are that:
  * At least 2 drug sets are represented in your drug sensitivity data frame
  * At least 3 samples (e.g., cell lines) are in the drug sensitivity data frame

Required inputs:
  * prediction.results: Dataframe containing enrichment results for
  mechanism-of-action (moa) pairs (i.e., results output by landscape or
  landscape_2d functions) based on predicted drug sensitivity scores.
  * og.results: Dataframe containing enrichment results for
  mechanism-of-action (moa) pairs (i.e., results output by landscape or
  landscape_2d functions) based on original drug sensitivity scores (i.e.,
  from a real drug screen).

Example:
1. Prepare drug.moa data frame with moa annotations for each drug
```{r}
# create list of drug names
Drug <- paste0("Drug_", seq(from = 1, to = 12))

# Drug is our default name for the column containing drug names
drug.moa <- as.data.frame(Drug)

# moa is our default name for the column containing drug moa set annotations
drug.moa$moa <- rep(paste("Set", LETTERS[seq(from = 1, to = 2)]), 6)
```

2. Perform MOA landscape for original data set
```{r}
## Step 2a: prepare drug sensitivity data frame to represent original data
# create list of sample names
Sample_ID <- seq(from = 1, to = 21)

og.drug.sensitivity <- as.data.frame(Sample_ID)

# give each drug values representative of AUC sensitivity scores
for(i in seq_len(length(Drug))){
  og.drug.sensitivity[,c(Drug[i])] <- rnorm(length(Sample_ID),
                                        mean = 0.83, sd = 0.166)
}

## Step 2b: perform moa landscape and store results
og.landscape <- MOAL::landscape(og.drug.sensitivity, drug.moa)
```

3. Perform MOA landscape for prediction data set
```{r}
## Step 3a: prepare drug sensitivity data frame to represent prediction data
prediction.drug.sensitivity <- as.data.frame(Sample_ID)

# give each drug values representative of AUC sensitivity scores
for(i in seq_len(length(Drug))){
  prediction.drug.sensitivity[,c(Drug[i])] <- rnorm(length(Sample_ID),
                                        mean = 0.83, sd = 0.166)
}

## Step 3b: perform moa landscape and store results
prediction.landscape <- MOAL::landscape(prediction.drug.sensitivity, drug.moa)
```

4. Compare MOA landscape results from prediction to original
```{r}
performance.test <- MOAL::performance(prediction.landscape$results,
                                og.landscape$results)
```

# Session Info
```{r}
sessionInfo()
```
